Developing Decarbonisation Pathways for Irish homes in changing TIMES

### 2021-08-19 manuscript.tex
Most warning gone and need to work on results next

\par  The expertise among energy policy researchers in Ireland combined with the global network of Energy Technology Systems Analysis Program (ETSAP) modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways. Previous versions of the Irish TIMES model were extracted from the Pan European TIMES model and then updated with specific national data and assumptions. The previous Irish TIMES models were used to provide informative reports on Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}.  
The introduction of new climate policies and targets combined with technological advancements, means there is a need for a new ESOM to be built to explore Ireland's new climate targets and sectoral five-year carbon budgets, the new ``TIMES Ireland Model" (TIM) was built in 2021 to address this need \textbf{(SOURCE)}.
\par TIMES is a bottom-up optimisation energy-environment model with perfect foresight which is used to analyse various levels of spatial, temporal and sectoral resolution. Exogenous technical, physical, and regulatory constraints can be applied and technologies have specified fuel types, efficiency, lifetime, availability, environmental characteristics, and fixed, variable and operation costs. TIMES uses a linear programming optimizer matrix in General Algebraic Modeling System (GAMS), to compute all calculations.

@@

The dwelling stock data is obtained from the SEAI BER Public database, which is firstly filtered, then scaled up using Central Statistics Office (CSO) data. The SEAI BER public database is Ireland's national Energy Performance Certificate (EPC) register, which complies with the Energy Performance of Buildings Directive 2018/844/EU (EPBD). The BERs are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$). 
\par Over half of Irish dwellings are captured in the BER database and every BER displays over 200 energy related variables. The DEAP model which is used to assign BER labels,  is outlined by Dineen et al.\cite{Dineen2015ImprovedSource} in Section \ref{Intro:ESOM}. In terms of space heating, DEAP assumes that each dwelling is single zoned, the heating season is from October to May inclusive, and the heating times are 7am to 9am and 5pm to 11pm ( 56 hours / week ) during the heating season. One assumption, the internal temperature assumption, results in a large error, where space and water heating are overestimated in DEAP. Space heating consumption is reduced by 26\% after internal temperature adjustments were applied in TIM, this brought the final energy consumption in line with the top-down SEAI Energy Balance. DEAP assumes that all living areas ( living room, sitting room, and possibly kitchen) in dwellings are heated to 21$^{\circ}$C and non-living areas ( bedroom, bathroom, hallways ) are heated to 18$^{\circ}$C. Through the analysis of previous residential modelling in Ireland this is not the case, as the internal temperatures are often much lower, Table \ref{Table:InternalTemps} outlines the residential internal temperature assumptions applied in TIM.

@@

Where $ {\text{\euro}}{}_{rf-rt,a}$ is the retrofit cost from pre-retrofit BER rating (rf) to post-retrofit rating (rt) for archetype (a). $Ali_{r,t}$ is the cost data from Ali et al. \cite{Ali2020ABuildings}, it contains all BER ratings and retrofit types but does not include an archetype, so an archetype factor (AF) is introduced, 0.9 for apartments, 1 for attached, and 1.1 for detached dwellings.  $TABULA_{rf-rt,a}$ is the average retrofit cost from pre-retrofit BER rating to post-retrofit BER rating for an archetype, data is sourced from TABULA \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland}.  However with only 62 samples in total not all retrofits are included. In this case, TABULA is ignored and Ali et al. is the only source of retrofit costs. The retrofit costs are cross-checked with the residential cost-optimal report \cite{AECOM2020ReportBuildings}. 
\par To simplify the model, the cost of retrofit per BER rating is not used in TIM. Instead, an average weighted cost per archetype and per retrofit is calculated.

@@

The previous section outlined the cost of retrofitting, and this section looks at the expected savings of retrofitting existing dwellings and building new dwellings.  
The annual heating costs per archetype and per BER rating are sourced from SEAI's indicative energy running costs and the internal temperature assumptions in Table \ref{Table:InternalTemps} is included to calculate the costs, which are outlined in Table \ref{Table:FuelCost}

@@

The technical bottom-up development approach of the residential sector is combined with other economic sectors to meet macroeconomic defined demands. The energy service demands in the residential sector are driven by the total dwelling stock for non-archetype energy services and apartment, attached or detached dwelling stock for archetype energy services. 
\par The residential stock projections up to 2040 are taken from Economic and Social Research Institute's (ESRI's) housing demand estimates \cite{Bergin2020RegionalLevel}, beyond 2040 the growth rate is extrapolated. However, archetpye specfic growth is limited or unavailable, for this reason a simple linear regression, as shown in Eq. \ref{Eq:RegArchetype} is used to calculate the share of an archetype in a given year (y).  This is graphically represented in Figure \ref{fig:DwellingStoc}. The growth rate of apartments is the highest and increasing up to 2040, while attached and detached dwellings are decreasing, indicating a preference for smaller dwellings to increase in the future. The obsolescence rate of the existing dwellings is 0.35\%.

@@

The demand for space and water heating is outlined in the previous section, and this section explains how water and space heating are achieved through technology and retrofits. While the demands are exogenous, the choice of technologies to at least satisfy the demand are endogenous and is captured in Equation \ref{Eq:Technology}.

@@

All space and water heating technologies are archetype specific and some technologies deliver both water and space heating, while other technologies can only deliver one type of heating, the list of available technologies is outlined in \ref{sec:appendixA4}. Retrofits are included as they satisfy the residential space heating demand, even though retrofits do not consume fuel similar to a PV system, from that perspective. 
Danish Energy Agency's Technology Data for Individual Heat Production dataset was used  was input to TIM \cite{JointResearchCentre2017Techno-economicEU}, this dataset is also used in the EU TIMES model, the dataset has high granularity and is continuously updated.

In TIM, the number of main heating systems equals the number of dwellings, this assumption ignores any impact of secondary system heating. One exception to the rule is stoves, which can only be used as secondary heating systems.

The base year heating energy consumption is derived from the BER database, but uses census data \cite{CentralStatisticsOffice2016CensusIreland} from the CSO dataset E1008 to calibrate the fuel used in heating systems per archetype.

@@

Lighting, appliances, and pumps \& fans are all fueled by electricity, whereas cooking can be fueled by electricity, gas or LPG. As shown in Figure \ref{fig:IrelandServiceDemands}, these energy services accounted for 20\% of energy consumption in 2018, however, this share of consumption is expected to increase. 
Refrigeration, cloth washing, cloth drying, and dish washing are under the appliances category and data is sourced from Topten \cite{Michel2016EnergyREPORT} and SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}. Cooking is also sourced from Eurostat's ``Energy consumption in households by type and end-used" excel datasheet \cite{Eurostat2021Disaggregatednrg_d_hhq} and cross-checked with SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}.
\par Appliances not included in this list are simply under the ``Other Appliances" category in TIM and these final values are cross-checked with POTEnCIA. Other Appliances include television, multimedia, and computer technology. Lighting and pumps \& fans are archetype specific energy services as the data is sourced from the BER database.  
Equation \ref{Eq:Technology} is also used to find the technology in the energy service demand, but unlike the heating energy service demands, there is little or no choice in different technologies (k) for some of the demand categories (i), therefore, the model is restricted in lighting, appliances and pumps \& fans. However, the cost and efficiency of each technology does determine the overall investment.

\subsubsection{User Constraints}
\label{LitR:Constraints}
The TIM model operates on the basis of perfect foresight, but
\textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel} for this reason ``User Constraints" are applied in the residential sector to better reflect reality. The User Constraints applied to the residential sector include:
\begin{itemize}
 \item One Space and Water Heating System per household ( stoves excluded ) 
 \item Maximum 60\% of existing buildings to be retrofitted by 2030, and 95\% by 2070. Maximum interpolated between 2030 and 2070. 
 \item A prerequisite requirement for electrical heat pump installation is applied to each archetype. The percentage of dwellings which requires a shallow or deep retrofit before a heat pump can be installed (  B2 rating for heat pump ready dwelling ). Hybrid heat pumps are not included in this constraint, as they can supply higher water temperatures without significantly reducing performance. 
 \begin{table}[htbp]
  \centering
  \footnotesize
  \caption{Prerequisite Heat Pump Requirement}
    \begin{tabular}{r|rrr}
    \hline
    \multicolumn{1}{l}{Requirement} & \multicolumn{1}{l}{Apartment} & \multicolumn{1}{l}{Attached } & \multicolumn{1}{l}{Detached} \\ \hline
    No Retrofit  & 24.5\%   & 14.3\%   & 12\% \\
    Shallow Retrofit  & 28.5\%   & 34.7\%  & 36.8\% \\
    Deep Retrofit  & 47\%   & 51\%  & 51.2\% \\
    \end{tabular}%
  \label{tab:HeatPumpConstraint}%
\end{table}%
 \item For space and water heating, each archetype has an upper gas growth constraint of 40\% higher than the base year value.
\item Maximum fuel share in cooking for gas and LPG is 40\% and 10\% respectively. Base year fuel share for Gas is 32\% and LPG is 1.5\%.
\item Maximum 10\% solar input to dual fuel heating technology
\item Maximum 30\% gas input to gas hybrid heating pumps
\end{itemize}

While other constraints are applied in different sectors, which affect the residential sector and vice versa, they are not listed in this paper. The growth constraints listed above are calculated using the following \cite{Loulou2016DocumentationI} : 
\begin{equation} 
\begin{split}
VAR\_CAP(t + 1) \leq (1 + G^{M(t+1)-M(t)}) \cdot VAR\_CAP(t)+K
\end{split}
\label{Eq:GrowthConstraint}
\end{equation}

Where $VAR\_CAP$ is the technology capacity in time period (t), G is the growth coefficient, defined as a new attribute of the technology and represents the maximum annual growth allowed for the capacity. The quantity $M(t+1)-M(t)$ is the number of years between the milestones of time periods t and t+1. The constant K is useful whenever the technology has no capacity initially, to allow the capacity to build over time.

\section{Results}
\label{Results}

>>>>

\par  The expertise among energy policy researchers in Ireland combined with the global network of Energy Technology Systems Analysis Program (ETSAP) modellers, makes TIMES the preferred ESOM to explore Ireland's decarbonisation pathways. Previous versions of the Irish TIMES model were extracted from the Pan European TIMES model and then updated with specific national data and assumptions. The previous Irish TIMES models were used to provide informative reports on Irish climate policy \cite{Gallachoir2012EPAModel,Deane2017Irish2,Gallachoir2020The3,Deane2013TechnicalIreland,IrishGovernment2017National2017}.  
The introduction of new climate policies and targets combined with technological advancements, means there is a need for a new ESOM to be built to explore Ireland's new climate targets and sectoral five-year carbon budgets, the new ``TIMES Ireland Model" (TIM) was built in 2021 to address this need \textbf{(SOURCE)}.
\par TIMES is a bottom-up optimisation energy-environment model with perfect foresight which is used to analyse various levels of spatial, temporal and sectoral resolution. Exogenous technical, physical, and regulatory constraints can be applied and technologies have specified fuel types, efficiency, lifetime, availability, environmental characteristics, and fixed, variable and operation costs. TIM is set up with regional options, either one national region or 26 county regions, only the annual results were explored in this study. TIMES uses a linear programming optimizer matrix in General Algebraic Modeling System (GAMS), to compute all calculations.

@@

The dwelling stock data is obtained from the SEAI BER Public database, which is firstly filtered, then scaled up using Central Statistics Office (CSO) data. The SEAI BER public database is Ireland's national Energy Performance Certificate (EPC) register, which complies with the Energy Performance of Buildings Directive 2018/844/EU (EPBD). The BERs are categorised based on expected energy consumption, with A1 representing the lowest expected energy consumption ($\leq 74 kWh/m^2/year$) and G representing the highest expected energy consumption ($ >450 kWh/m^2/year$). 
\par Over half of Irish dwellings are captured in the BER database and every BER displays over 200 energy related variables. The DEAP model which is used to assign BER labels,  is outlined by \citet{Dineen2015ImprovedSource} in Section \ref{Intro:ESOM}. In terms of space heating, DEAP assumes that each dwelling is single zoned, the heating times are 7am to 9am and 5pm to 11pm ( 56 hours / week ) during the heating season, which is from October to May inclusive and assumes monthly external temperatures, the average external temperature for the heating season is 7.6$^{\circ}$C. One assumption, the internal temperature assumption, results in a large error, where space and water heating are overestimated in DEAP. Space heating consumption is reduced by 26\% after internal temperature adjustments were applied in TIM, this brought the final energy consumption in line with the top-down SEAI Energy Balance. DEAP assumes that all living areas ( living room, sitting room, and possibly kitchen) in dwellings are heated to 21$^{\circ}$C and non-living areas ( bedroom, bathroom, hallways ) are heated to 18$^{\circ}$C. Through the analysis of previous residential modelling in Ireland this is not the case, as the internal temperatures are often much lower, Table \ref{Table:InternalTemps} outlines the residential internal temperature assumptions applied in TIM.

@@

Where $ {\text{\euro}}{}_{rf-rt,a}$ is the retrofit cost from pre-retrofit BER rating (rf) to post-retrofit rating (rt) for archetype (a). $Ali_{r,t}$ is the cost data from  \citet{Ali2020ABuildings}, it contains all BER ratings and retrofit types but does not include an archetype, so an archetype factor (AF) is introduced, 0.9 for apartments, 1 for attached, and 1.1 for detached dwellings.  $TABULA_{rf-rt,a}$ is the average retrofit cost from pre-retrofit BER rating to post-retrofit BER rating for an archetype, data is sourced from TABULA \cite{IrishTABULANationalAdvisoryGroup2014BuildingIreland}.  However with only 62 samples in total not all retrofits are included. In this case, TABULA is ignored and \citet{Ali2020ABuildings} is the only source of retrofit costs. The retrofit costs are cross-checked with the residential cost-optimal report \cite{AECOM2020ReportBuildings}. 
\par To simplify the model, the cost of retrofit per BER rating is not used in TIM. Instead, an average weighted cost per archetype and per retrofit is calculated.

@@

The previous section outlined the cost of retrofitting, and this section looks at the expected savings of retrofitting existing dwellings and building regulations in new dwellings.  
The annual heating costs per archetype and per BER rating are sourced from SEAI's indicative energy running costs. The internal temperature assumptions in Table \ref{Table:InternalTemps} are included to calculate the costs, which are outlined in Table \ref{Table:FuelCost}

@@

The  technical  bottom-up  development  approach  of  the  residential  sector  is  combined with other economic sectors to meet macroeconomic defined demands. The energy service demands in the residential sector are driven by the total dwelling stock for non-archetype energy services and apartment, attached or detached for archetype energy services. The residential stock projections up to 2040 are taken from ESRI's housing demand estimates \cite{Bergin2020RegionalLevel}, beyond 2040 the growth rate is extrapolated. However, the archetype specific growth is limited or unavailable, for this reason a simple linear regression, as shown in Eq. \ref{Eq:RegArchetype} is used to calculate the share of an archetype in a given year (y).  This is graphically represented in Figure \ref{fig:DwellingStoc}. The growth rate of apartments is the highest and increasing up to 2040, while the growth in attached and detached dwellings are lagging behind. The obsolescence rate of the existing dwellings is taken as 0.35\%.

@@

The demand for space and water heating is outlined in the previous section, and this section explains how water and space heating are achieved through technology and retrofits. While the demands are exogenous, the choice of technologies to at least satisfy the demand are endogenous and is captured in Eq. \ref{Eq:Technology}.

@@

All space and water heating technologies are archetype specific and some technologies deliver both water and space heating, while other technologies can only deliver one type of heating, the list of available technologies is outlined in \ref{sec:appendixA4}. Retrofits are included as they satisfy the residential space heating demand, even though retrofits do not consume fuel similar to a solar water heating system, from that perspective. 
Danish Energy Agency's Technology Data for Individual Heat Production dataset was used  was input to TIM \cite{JointResearchCentre2017Techno-economicEU}, this dataset is also used in the EU TIMES model, the dataset has high granularity and is continuously updated.

In TIM, the number of main heating systems equals the number of dwellings, this assumption ignores any impact of secondary system heating. One exception to the rule is stoves are considered as secondary heating systems only.  The base year heating energy consumption is derived from the BER database, but uses census data \cite{CentralStatisticsOffice2016CensusIreland} from the CSO dataset E1008 to calibrate the fuel used in heating systems per archetype.

@@

As shown in Figure \ref{fig:IrelandServiceDemands}, these energy services accounted for 20\% of energy consumption in 2018, however, this share of consumption is expected to increase. 
Refrigeration, cloth washing, cloth drying, and dish washing are under the appliances category and data is sourced from Topten \cite{Michel2016EnergyREPORT} and SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}. Cooking data is sourced from Eurostat's ``Energy consumption in households by type and end-used"  \cite{Eurostat2021Disaggregatednrg_d_hhq} and cross-checked with SEAI \cite{SustainableEnergyAuthorityofIreland2018EnergySector}.
\par Appliances not included in this list are simply under the ``Other Appliances" category in TIM. Other Appliances include television, multimedia, outdoor appliances, and computer technology. Lighting and pumps \& fans are archetype specific energy services as the data is sourced from the BER database.  
\par Equation \ref{Eq:Technology} is also used to optimal choose the technology in the energy service demand, but unlike in the heating energy service demands, there is little or no choice in different technologies (k) for some of the demand categories (i), therefore, the model is restricted in lighting, appliances and pumps \& fans. However, the cost and efficiency of each technology does determine the overall investment.

\subsubsection{User Constraints}
\label{LitR:Constraints}
The TIM model operates on the basis of perfect foresight, but
\textit{``in reality, the behaviours of consumers is not always economically rational"} \cite{Li2018IncorporatingModel} for this reason ``User Constraints" are applied in the residential sector to better reflect reality. The User Constraints applied to the residential sector include:
\begin{enumerate}[1.]
 \item Only one space and water heating system per household ( stoves excluded ) 
 \item Maximum 60\% of existing buildings to be retrofitted by 2030, and 95\% by 2070. Maximum interpolated between 2030 and 2070. 
 \item A prerequisite requirement for electrical heat pump installation is applied to each archetype. The percentage of dwellings which requires a shallow or deep retrofit before a heat pump can be installed (  B2 rating for heat pump ready dwelling ). Hybrid heat pumps are not included in this constraint, as they can supply higher water temperatures without significantly reducing performance. 
 \begin{table}[htbp]
  \centering
  \footnotesize
  \caption{Prerequisite Heat Pump Requirement}
    \begin{tabular}{r|rrr}
    \hline
    \multicolumn{1}{l}{Requirement} & \multicolumn{1}{l}{Apartment} & \multicolumn{1}{l}{Attached } & \multicolumn{1}{l}{Detached} \\ \hline
    No Retrofit  & 24.5\%   & 14.3\%   & 12\% \\
    Shallow Retrofit  & 28.5\%   & 34.7\%  & 36.8\% \\
    Deep Retrofit  & 47\%   & 51\%  & 51.2\% \\
    \end{tabular}%
  \label{tab:HeatPumpConstraint}%
\end{table}%
 \item For space and water heating, each archetype has an upper gas growth constraint of 40\% higher than the base year value.
\item Maximum fuel share in cooking for gas and LPG is 40\% and 10\% respectively. Base year fuel share for Gas is 32\% and LPG is 1.5\%.
\item Maximum 10\% solar input to dual fuel heating technology
\item Maximum 30\% gas input to gas hybrid heating pumps
\end{enumerate}

While other constraints are applied in different sectors, which affect the residential sector and vice versa, they are not listed in this paper. The growth constraints listed above are calculated using the following \cite{Loulou2016DocumentationI} : 
\begin{equation} 
\begin{split}
VAR\_CAP(t + 1) \leq (1 + G^{M(t+1)-M(t)}) \cdot VAR\_CAP(t)
\end{split}
\label{Eq:GrowthConstraint}
\end{equation}

Where $VAR\_CAP$ is the technology capacity in time period (t), G is the growth coefficient, defined as a new attribute of the technology and represents the maximum annual growth allowed for the capacity. The quantity $M(t+1)-M(t)$ is the number of years between the milestones of time periods t and t+1.

%PICKUPHERE
\section{Results}
\label{Results}
